Apples and Oranges in Adolescent Depression Treatments.

### 2024-01-03 apples_and_oranges.qmd
Few small changes - checking df_long_for_metan and added method re imputation of missing SDs and means

---
title: Apples and Oranges in Adolescent Depression Treatments.
subtitle: A conceptual and quantiative re-appraisal of control arms in clinical trials.
author:
  - name: Argyris Stringaris
    email: a.stringaris@ucl.ac.uk
    affiliations: 
        - id: ucl_1
          name: University College London
          department: Divisions of Psychiatry and Psychology and Language Science
          address: 1-19 Torrington Place
          city: London, UK
          state: 
          postal-code: WC1E 7HB
          country: United Kingdom
    attributes:
        corresponding: true
    note: 
  - name: Charlotte Burman
    email: c.burman@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
  - name: Dayna Bhudia
    email: d.bhudia@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
  - name: Carmen Moreno
    email: cmoreno@hggm.es
    affiliations:
        - id: HGM
          name: Hospital Gregorio Marañón
          department: Department of Psychiatry
          address: 46 C. del Dr. Esquerdo
          city: Madrid
          state:
          postal-code: 28007
          country: Spain
  - name: Samuele Cortese
    email: samuele.cortese@soton.ac.uk
    affiliations:
        - id: soton
          name: University of Southampton
          department: Centre for Innovation in Mental Health (CIMH), School of Psychology
          address: Highfield Campus, Building 44
          city: Southampton
          state:
          postal-code: SO17 1BJ
          country: United Kingdom
  - name: Georgina Krebs
    email: g.krebs@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note: 
abstract: |
  To follow
keywords: 
  - depression
  - adolescents
  - antidepressants
  - psychotherapy
date: last-modified
bibliography: emotion_concepts_paper.bib
format: 
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super

@@

Clinicians across medicine and other health professions are faced with choices betweent different treatment modalities for their patients, be it between surgical or medical (drug-based) treatments or, in the case of adolescent depression, between psychotherapy and antidepressant medication. To make informed choices, clinicians rely on the extant evidence, frequently summarised in the form of national guidelines or other treatment recommendations. Such a comparison is inherent in the internationally influential UK National Institute of Clinical Excellence (NICE) guidelines for adolescent depression which recommend that youth be first offered psychological therapy over medication for most presentations of depression. Such comparisons between the etwo treatment modalities are inherent in other recommendations, e.g. the practice parameter issued by the American Academy of Child and Adolescent Psychiatry or the Royal Australian and New Zealand College of Psychiatrists.

These recommendations are not based on head to head trials between modalities; only one head-to-head trial exists in adolescent depression Instead, they rely on the comparison of the results of randomised controlled trials (RCTs) of each modality against RCTs of the other. In other words, it rests on a comparison of comparisons.

@@

NOTE: **Charlotte** to please check code lines 986 to 1121
NOTE: **Charlotte** please check that the studies I have in the *df_long_for_metan* dataframe makes sense and correspond to what you would expect. In the beginning of the results below, I explain in detail what is missing where.

@@

where, $Mean_{t_{2}}$ and $Mean_{t_{1}}$ refer to the means of the main outcome score at the end and beginning of the study respectively and $SD_{t_{2}}$ and $SD_{t_{1}}$ to the respective standard deviations. NOTE **Charlotte**, please add note here with what we did to impute SDs where missing, e.g. using the pre to impute the post etc. In general state here the rules we used citing the Cochrane book.

>>>>

---
title: Apples and Oranges in Adolescent Depression Treatments.
subtitle: A conceptual and quantiative re-appraisal of control arms in clinical trials.
author:
  - name: Argyris Stringaris
    email: a.stringaris@ucl.ac.uk
    affiliations: 
        - id: ucl_1
          name: University College London
          department: Divisions of Psychiatry and Psychology and Language Science
          address: 1-19 Torrington Place
          city: London, UK
          state: 
          postal-code: WC1E 7HB
          country: United Kingdom
    attributes:
        corresponding: true
    note: 
  - name: Charlotte Burman
    email: charlotte.burman@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
  - name: Dayna Bhudia
    email: dayna.bhudia@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
  - name: Carmen Moreno
    email: cmoreno@hggm.es
    affiliations:
        - id: HGM
          name: Hospital Gregorio Marañón
          department: Department of Psychiatry
          address: 46 C. del Dr. Esquerdo
          city: Madrid
          state:
          postal-code: 28007
          country: Spain
  - name: Samuele Cortese
    email: samuele.cortese@soton.ac.uk
    affiliations:
        - id: soton
          name: University of Southampton
          department: Centre for Innovation in Mental Health (CIMH), School of Psychology
          address: Highfield Campus, Building 44
          city: Southampton
          state:
          postal-code: SO17 1BJ
          country: United Kingdom
  - name: Georgina Krebs
    email: g.krebs@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note: 
abstract: |
  To follow
keywords: 
  - depression
  - adolescents
  - antidepressants
  - psychotherapy
date: last-modified
bibliography: emotion_concepts_paper.bib
format: 
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super

@@

Clinicians across medicine and other health professions are faced with choices between different treatment modalities for their patients, be it between surgical or medical (drug-based) treatments or, in the case of adolescent depression, between psychotherapy and antidepressant medication. To make informed choices, clinicians rely on the extant evidence, frequently summarised in the form of national guidelines or other treatment recommendations. Such a comparison is inherent in the internationally influential UK National Institute of Clinical Excellence (NICE) guidelines for adolescent depression which recommend that youth be first offered psychological therapy over medication for most presentations of depression. Such comparisons between the two treatment modalities are inherent in other recommendations, e.g. the practice parameter issued by the American Academy of Child and Adolescent Psychiatry or the Royal Australian and New Zealand College of Psychiatrists.

These recommendations are not based on head to head trials between modalities; only one head-to-head trial exists in adolescent depression. Instead, they rely on the comparison of the results of randomised controlled trials (RCTs) of each modality against RCTs of the other. In other words, it rests on a comparison of comparisons.

@@

NOTE: **Charlotte** to please check code lines 986 to 1121. **CB**: In progress.
NOTE: **Charlotte** please check that the studies I have in the *df_long_for_metan* dataframe makes sense and correspond to what you would expect. In the beginning of the results below, I explain in detail what is missing where. **CB**: I've checked and there are fewer psy active conditions that I expected - it seems some of the trials with multiple active arms have been filtered out. Also, for psy the no of studies with missing Cohens d for controls should probably be considered 6 (not 7) as it is absent for March/TADS because we removed it (didn't want placebo to be considered as a psy control) rather than there being missing data. The numbers look right for med studies.

@@

where, $Mean_{t_{2}}$ and $Mean_{t_{1}}$ refer to the means of the main outcome score at the end and beginning of the study respectively and $SD_{t_{2}}$ and $SD_{t_{1}}$ to the respective standard deviations. NOTE **Charlotte**, please add note here with what we did to impute SDs where missing, e.g. using the pre to impute the post etc. In general state here the rules we used citing the Cochrane book. **CB** Where individual studies did not report all data required to calculate the SMD, we imputed missing data according to the methods recommended by X (citation), in the following order. If a study reported the standard error of the mean, the SD was obtained simply by multiplying the SE by the square root of the sample size. For conditions where the SD was missing at one time point, the baseline SD was substituted by the post-test SD, and vice versa. If the SD was not available at either time point, an average SD was calculated by taking the mean of the SDs of all trials of the same type (i.e. psy or med) using the same instrument. These average SDs were calculated for each time point and for each arm (control and active). Where there were missing means at either baseline or post-test, missing values were calculated using mean change scores, preferring the change scores reported in the paper itself, though where this was unavailable, using the change scores reported in the dataset from Cipriani's meta-analysis (citation).
